### 2023-03-01 calculus.tex
Rename marking calculus to marked lambda calculus

\section{The Marking Calculus}
\label{sec:calculus}

@@

The key contribution of this section is the \emph{marking calculus}, a formal calculus for type
error localization in bidirectional systems. Namely, the calculus provides a judgmental framework
with a unified theory for the design of typing semantics that support ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
lambda calculus, extending it with binary product types in \cref{sec:calculus-products}. All rules
and theorems may be found in the supplementary appendix, alongside a complete mechanization in the
Agda proof assistant, which is discussed briefly in \cref{sec:calculus-agda}. With products, we then
consider how error localization is complicated by patterned let expressions in
\cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
to more complex typing features, such as System F-style polymorphism.

@@

We now formally introduce the \emph{marking calculus} through its application to a gradually typed
lambda calculus extended with numbers and booleans. Consider the syntax for such a language, given
as $\EMName$ in \cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and
boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
$\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$
correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
$\TUnknown$ is the unknown type from gradual type theory. $\EEHole$ denotes an \emph{expression
hole}, which are used to model syntactically incomplete programs---but are not critical within the
calculus.

@@

The \emph{marking calculus} builds atop these typing semantics into a two-layer system:
%
\begin{itemize}
  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

@@

  \judgment{
    \ctxSynType{\ctx}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\\\
    \subsumable{\ECMV}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECMV}}{\TMV}
  }{MAInconsistentTypes}
\end{mathpar}
%
These rules complete the development of the marking calculus for the gradually typed lambda
calculus.

@@

We now demonstrate how one might extend or adapt the marking calculus for new typing features by
adding binary product types. First, in \cref{fig:calculus-product-judgments}, we extend the syntax
of types and the accompanying notions of consistency and join, which closely mirror the definitions
for arrow types.
%
\[\begin{array}{rrcl}
  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TPair{\TMV}{\TMV}
\end{array}\]

@@

\section{The Marking Calculus in Structured Editing}
\label{sec:calculus-structured-editing}

@@

By applying the marking calculus to Hazelnut, the action calculus may be \emph{wholly decoupled}
from the typing semantics. Under this new system, the action calculus is concerned only with the
manipulation of syntax, and the marking procedure yields statically meaningful terms with error
holes inserted at the appropriate positions. Furthermore, marking solves the problem of non-local
hole fixes; actions may be performed anywhere (as long as they are of the correct sort), and marking
totality ensures that they give meaningful terms.
% TODO this is not strictly true, only for 'wrapping' actions

>>>>

\section{The Marked Lambda Calculus}
\label{sec:calculus}

@@

The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
type error localization in bidirectional systems. Namely, the calculus provides a judgmental
framework with a unified theory for the design of typing semantics that support ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
lambda calculus, extending it with binary product types in \cref{sec:calculus-products}. All rules
and theorems may be found in the supplementary appendix, alongside a complete mechanization in the
Agda proof assistant, which is discussed briefly in \cref{sec:calculus-agda}. With products, we then
consider how error localization is complicated by patterned let expressions in
\cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
to more complex typing features, such as System F-style polymorphism.

@@

We now formally introduce the marked lambda calculus through its application to a gradually typed
lambda calculus extended with numbers and booleans. Consider the syntax for such a language, given
as $\EMName$ in \cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and
boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
$\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$
correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
$\TUnknown$ is the unknown type from gradual type theory. $\EEHole$ denotes an \emph{expression
hole}, which are used to model syntactically incomplete programs---but are not critical within the
calculus.

@@

The marked lambda calculus builds atop these typing semantics into a two-layer system:
%
\begin{itemize}
  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

@@

  \judgment{
    \ctxSynType{\ctx}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\\\
    \subsumable{\ECMV}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECMV}}{\TMV}
  }{MAInconsistentTypes}
\end{mathpar}
%
These rules complete the development of the marked lambda calculus for the gradually typed lambda
calculus.

@@

We now demonstrate how one might extend or adapt the marked lambda calculus for new typing features
by adding binary product types. First, in \cref{fig:calculus-product-judgments}, we extend the
syntax of types and the accompanying notions of consistency and join, which closely mirror the
definitions for arrow types.
%
\[\begin{array}{rrcl}
  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TPair{\TMV}{\TMV}
\end{array}\]

@@

\section{The Marked Lambda in Structured Editing}
\label{sec:calculus-structured-editing}

@@

By applying the marked lambda calculus to Hazelnut, the action calculus may be \emph{wholly decoupled}
from the typing semantics. Under this new system, the action calculus is concerned only with the
manipulation of syntax, and the marking procedure yields statically meaningful terms with error
holes inserted at the appropriate positions. Furthermore, marking solves the problem of non-local
hole fixes; actions may be performed anywhere (as long as they are of the correct sort), and marking
totality ensures that they give meaningful terms.
% TODO this is not strictly true, only for 'wrapping' actions
